Tree establishment in models of forest dynamics: a key priority for further research

### 2023-03-12 manuscript/manuscriptRW.Rmd
added proper PROCLIAS funding text in the manuscript

When evaluating the representation of ingrowth levels along key
gradients of light availability (basal area), temperature (degree-days)
and soil moisture (climatic water balance), distinct features emerge:
the models reproduced the effect of basal area in magnitude and patterns
(Figure \@ref(fig:nvTrend710BA)) better than the effects of the climatic
gradients, where they featured varying effects (Figures S1 and S2).

In the observed data, ingrowth levels decreased clearly with increasing
stand basal area (Figure \@ref(fig:nvTrend710BA)). This trend was
captured by the Landis-II model only, albeit at much lower values of
basal area than in the observations. The other models featured
distinctly different trends, such as (1) an increase in ingrowth levels
with increasing total basal area followed by the absence of ingrowth at
high values of basal area (ForCEEPS, ForCEEPS(f), FORMIND, TreeMig and
LPJ-GUESS), (2) almost constant ingrowth levels with basal area (ForClim
11, xComp, LandClim and aDGVM2), or (3) an increase of ingrowth up to a
certain value of stand basal area followed by a decrease at even higher
values of stand basal area, with model-specific thresholds (4C, ForClim
1, PICUS, SIBYLA and iLand).

The models did not match the observed ingrowth patterns across gradients
of soil moisture and temperature (Figures S1 and S2), and neither was
there a clear pattern across models. Regarding the soil moisture
gradient (Figure S1), a group of models featured decreasing ingrowth
with increasing soil moisture (4C, ForClim 1, ForClim11, SIBYLA, xComp,
PICUS, LandClim, TreeMig and iLand), while a few models showed the
opposite trend (ForCEEPS(f), FORMIND and LPJ-GUESS). Across the
temperature gradient (Figure S2), all the models that featured a
decrease in ingrowth levels with increasing soil moisture showed the
opposite trend (i.e., more ingrowth with increasing temperature), except
for 4C. The other models did not feature a clear pattern.

```{r nvTrend710BAlightTemp, echo=FALSE, fig.cap="Mean ingrowth values across the 200 samples per site, for the 200 sites against gradients of (a) total basal area; (b) climatic water balance; (c) seasonal degree sum. The values were split into ten bins; the red lines represent a GAM model showing the trend in the observed data. The same graphs with y-axis scaled by model as availalbe in Figure S", fig.show = 'hold', out.width="49%",fig.ncol = 2, fig.subcap=c('', '', ''), fig.show='hold', fig.align='left'}

   knitr::include_graphics(c(here::here("figures", "envTrend7Totba_500.png"),
                              here::here("figures", "envTrend7WB_500.png"),
                              here::here("figures", "envTrend7DDS_500.png")))
```

Silver fir (*Abies alba*), beech (*Fagus sylvatica*), Scots pine (*Pinus
sylvestris*) and oak (*Quercus* spp.) were the species observed most
frequently in the EuFoRIa data. Most models captured well the share of
basal area in the ingrowth of these five species compared to the
observed data along the two major gradients of climate (Figure
\@ref(fig:diff), cf. Figure S7). However, the models differed
considerably in the way in which they simulated their climatic niche.

@@

Funding of the workshop by COST (Action No CA19139) and ETH Zurich
(Grant ETH-35 18-1).

>>>>

When evaluating the representation of ingrowth levels along key
gradients of light availability (basal area), temperature (degree-days)
and soil moisture (climatic water balance), distinct features emerge:
the models reproduced the effect of basal area in magnitude and patterns
(Figure \@ref(fig:nvTrend710BAlightTemp) and S7) better than the effects
of the climatic gradients, where they featured varying effects.

In the observed data, ingrowth levels decreased clearly with increasing
stand basal area (Figure \@ref(fig:nvTrend710BAlightTemp)). This trend
was captured by the Landis-II model only, albeit at much lower values of
basal area than in the observations. The other models featured
distinctly different trends, such as (1) an increase in ingrowth levels
with increasing total basal area followed by the absence of ingrowth at
high values of basal area (ForCEEPS, ForCEEPS(f), FORMIND, TreeMig and
LPJ-GUESS), (2) almost constant ingrowth levels with basal area (ForClim
11, xComp, LandClim and aDGVM2), or (3) an increase of ingrowth up to a
certain value of stand basal area followed by a decrease at even higher
values of stand basal area, with model-specific thresholds (4C, ForClim
1, PICUS, SIBYLA and iLand).

The observed ingrowth did not change much across the climatic water
balance gradient, and showed a slight increase with temperature. The
models did not match the observed ingrowth patterns across gradients of
soil moisture and temperature (Figure
\@ref(fig:)\@ref(fig:nvTrend710BAlightTemp) and S7), and neither was
there a clear pattern across models. Regarding the soil moisture
gradient, a group of models featured decreasing ingrowth with increasing
soil moisture (4C, ForClim 1, ForClim11, SIBYLA, xComp, PICUS, LandClim,
TreeMig and iLand), while a few models showed the opposite trend
(ForCEEPS(f), FORMIND and LPJ-GUESS). Across the temperature gradient,
all the models that featured a decrease in ingrowth levels with
increasing soil moisture showed the opposite trend (i.e., more ingrowth
with increasing temperature), except for 4C. The other models did not
feature a clear pattern.

```{r nvTrend710BAlightTemp, echo=FALSE, fig.cap="Mean ingrowth values across the 200 samples per site with y-axis scaled by model, for the 200 sites against gradients of (a) total basal area; (b) climatic water balance; (c) seasonal degree sum. The values were split into ten bins; the red lines represent a GAM model showing the trend in the observed data. For comparing the magnitude of recruitment across models the same graphs with equal y-axis are availalbe in Figure S7", fig.show = 'hold', out.width="49%",fig.ncol = 2, fig.subcap=c('', '', ''), fig.show='hold', fig.align='left'}

   knitr::include_graphics(c(here::here("figures", "envTrend7Totba.png"),
                              here::here("figures", "envTrend7WB.png"),
                              here::here("figures", "envTrend7DDS.png")))
```

Silver fir (*Abies alba*), beech (*Fagus sylvatica*), Scots pine (*Pinus
sylvestris*) and oak (*Quercus* spp.) were the species observed most
frequently in the EuFoRIa data. Most models captured well the share of
basal area in the ingrowth of these five species compared to the
observed data along the two major gradients of climate (Figure
\@ref(fig:diff), cf. Figure S8). However, the models differed
considerably in the way in which they simulated their climatic niche.

@@

Funding of the workshop by COST Action CA19139 PROCLIAS (PROcess-based
models for CLimate Impact Attribution across Sectors), supported by COST
(European Cooperation in Science and
Technology; [https://www.cost.eu](https://www.cost.eu/)) and ETH Zurich
(Grant ETH-35 18-1).
